---
title: "CAZ_prescription_trends"
---

## Overview

This section presents a comprehensive analysis of prescription trends in areas affected by Clean Air Zones (CAZ). The analysis focuses primarily on Short-Acting Beta Agonists (SABA) inhalers, which are commonly prescribed for asthma and other respiratory conditions.

By examining prescription patterns before and after CAZ implementation, and comparing areas within CAZ boundaries to surrounding areas, we can assess whether air quality improvements are associated with changes in respiratory medication use. A reduction in SABA prescriptions could indicate improved respiratory health outcomes in the population.

The analysis incorporates:

-   **Temporal trends**: Changes in prescription rates over time
-   **Spatial patterns**: Differences between CAZ and non-CAZ areas
-   **Practice-level analysis**: Variations across individual GP practices
-   **Population adjustments**: Accounting for patient registration numbers

This comprehensive approach helps to identify potential causal relationships between air quality interventions and health outcomes while controlling for confounding factors.

## Setup

### Package Installation

Loading required packages for data analysis and visualisation.

```{r}
#| label: packages
#| message: false

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("remotes")) {
  install.packages("remotes")
}
pkgs = c(
  "sf",
  "tidyverse",
  "here",
  "tmap",
  "data.table"
)

remotes::install_cran(pkgs)
sapply(pkgs, require, character.only = TRUE)

dir.create(
  file.path(
    here(),
    "qmds_ghpages/figures"
  ),
  showWarnings = FALSE
)

```

## Loading data

### GP practices

Loading spatial and administrative data for GP practices and CAZ areas.

```{r}
#| label: read-data-pract
subicb_boundaries <- st_read(
  file.path(
    here(),
    "data_raw",
    "subicb_boundaries.gpkg"
  ),
  quiet = TRUE
)
pract_classed <- st_read(file.path(here(), "data_raw", "CAZ_practices.gpkg"),
 quiet = TRUE)
all_practices <- st_read(
  file.path(here(), "data_raw", "all_practices_urban_rural.gpkg"),
  quiet = TRUE
)
CAZ_lst <- read_csv("data_raw/CAZ_list.csv")
```

### Patient data

Loading patient registration data for calculating prescription rates.

```{r}
#| label: read-data-patients

patient_num_df <- read_csv("data_raw/practice_patients.csv")

patient_num_df$month <- match(patient_num_df$month, tolower(month.name))
```

### Prescription data

Loading SABA prescription data for respiratory health analysis.

```{r}
#| label: read-data-prescr-saba

all_saba_raw <- read_csv(file = file.path(here(), "data_raw", "all_saba.csv"))
```

Loading comprehensive BNF prescription data for broader medication analysis.

```{r}
#| label: read-data-prescr-main

get_all_data <- function(group) {
  lst_files <- list.files(
    "data_raw/bnf_all/",
    pattern = paste0("^\\d{", group, "}_\\d{4}"),
    full.names = TRUE
  )
  lapply(
    lst_files,
    \(i) {
      fread(i)[,
        bnf := str_extract(
          pattern = paste0("^\\d{", group, "}"),
          string = basename(i)
        )
      ]
    }
  ) |>
    rbindlist()
}

all_data <- get_all_data(2)
```

Loading data for Montelukast

```{r}
get_code_data <- function(code) {
  lst_files <- list.files(
    "data_raw/bnf_all/",
    pattern = code,
    full.names = TRUE
  )
  lapply(
    lst_files,
    \(i) {
      fread(i)[, bnf := str_extract(pattern = code, string = basename(i))]
    }
  ) |>
    rbindlist()
}
all_mtlkt_data <- get_code_data(code = "0303020G0")
all_antd_data <- get_code_data(code = "0403")
all_anx_data <- get_code_data(code = "040102")
all_meth_data <- get_code_data(code = "0603020S0")

# Load new cardiovascular indicators
all_thiazide_data <- get_code_data(code = "020201")
all_ace_data <- get_code_data(code = "0205051")
all_arb_data <- get_code_data(code = "020505")
all_bb_data <- get_code_data(code = "0204")
all_statin_data <- get_code_data(code = "0212")
```


## Preparing data

### Classifying practices between CAZ and non-CAZ (except London)

Creating location-based classifications for GP practices relative to CAZ boundaries.

```{r}
CAZ_practices <- pract_classed |>
  mutate(
    location_class = case_when(
      buffer_km == 0 ~ "within CAZ",
      buffer_km <= 1 ~ "within 1km",
      buffer_km <= 5 ~ "within 5km",
      TRUE ~ "out CAZ"
    )
  ) |>
  mutate(
    location_class = factor(
      location_class,
      levels = c(
        "within CAZ",
        "within 1km",
        "within 5km",
        "out CAZ",
        "Control"
      ),
      ordered = T
    )
  )

all_valid_subicb_codes <- subicb_boundaries |>
  st_drop_geometry() |>
  filter(str_detect(name, "LONDON", negate = TRUE)) |>
  pull(code)

London_subicb_geom_filter <- subicb_boundaries |>
  filter(str_detect(name, "LONDON")) |>
  st_union() |>
  st_buffer(20e3)


all_other_codes <- all_practices |>
  st_drop_geometry() |>
  filter(
    par_code %in% all_valid_subicb_codes,
    RUC21NM_adj %in%
      c(
        "Urban: within major town or city"
      ),
    !code %in% CAZ_practices$code
  ) |>
  pull(code)

```

A visualisation of the practices that will be included in the control

```{r}
all_practices |>
  filter(code %in% all_other_codes) |>
  qtm(fill = "RUC21NM_adj")
```

### Subsetting SABA data

Filtering SABA prescription data for CAZ and control areas with patient population adjustments.

```{r}

CAZ_SABA <- all_saba_raw |>
  filter(org_id %in% CAZ_practices$code) |>
  mutate(month = month(date), year = year(date)) |>
  left_join(
    patient_num_df |> select(CODE, NUMBER_OF_PATIENTS, month, year),
    by = c("org_id" = "CODE", "month", "year")
  )


noCAZ_SABA <- all_saba_raw |>
  filter(!org_id %in% CAZ_practices$code, org_id %in% all_other_codes) |>
  mutate(month = month(date), year = year(date)) |>
  left_join(patient_num_df, by = c("org_id" = "CODE", "month", "year")) |>
  mutate(location_class = "Control") |>
  mutate(
    location_class = factor(
      location_class,
      levels = c(
        "within CAZ",
        "within 1km",
        "within 5km",
        "out CAZ",
        "Control"
      ),
      ordered = T
    )
  )

```

### Data Preparation Helper

```{r}
# Define a helper function to prepare prescription data by classifying practice locations
prepare_med_data <- function(raw_data) {
  caz <- raw_data[row_id %in% CAZ_practices$code, ][, `:=`(
    month = month(date),
    year = year(date)
  )] |>
    left_join(
      patient_num_df |> select(CODE, NUMBER_OF_PATIENTS, month, year),
      by = c("row_id" = "CODE", "month", "year")
    )
  
  nocaz <- raw_data[
    !row_id %in% CAZ_practices$code & row_id %in% all_other_codes,
  ][, `:=`(month = month(date), year = year(date))] |>
    left_join(
      patient_num_df |> select(CODE, NUMBER_OF_PATIENTS, month, year),
      by = c("row_id" = "CODE", "month", "year")
    ) |>
    mutate(location_class = "Control") |>
    mutate(
      location_class = factor(
        location_class,
        levels = c("within CAZ", "within 1km", "within 5km", "out CAZ", "Control"),
        ordered = TRUE
      )
    )
  
  list(caz = caz, nocaz = nocaz)
}
```

### Subsetting data of all indicators

```{r}
res_bnf <- prepare_med_data(all_data)
CAZ_bnf <- res_bnf$caz
noCAZ_bnf <- res_bnf$nocaz

res_mtlkst <- prepare_med_data(all_mtlkt_data)
CAZ_mtlkst <- res_mtlkst$caz
noCAZ_mtlkst <- res_mtlkst$nocaz

res_antd <- prepare_med_data(all_antd_data)
CAZ_antd <- res_antd$caz
noCAZ_antd <- res_antd$nocaz

res_anx <- prepare_med_data(all_anx_data)
CAZ_anx <- res_anx$caz
noCAZ_anx <- res_anx$nocaz

res_meth <- prepare_med_data(all_meth_data)
CAZ_meth <- res_meth$caz
noCAZ_meth <- res_meth$nocaz

# Subset for new cardiovascular indicators
res_thiazide <- prepare_med_data(all_thiazide_data)
CAZ_thiazide <- res_thiazide$caz
noCAZ_thiazide <- res_thiazide$nocaz

res_ace <- prepare_med_data(all_ace_data)
CAZ_ace <- res_ace$caz
noCAZ_ace <- res_ace$nocaz

res_arb <- prepare_med_data(all_arb_data)
CAZ_arb <- res_arb$caz
noCAZ_arb <- res_arb$nocaz

res_bb <- prepare_med_data(all_bb_data)
CAZ_bb <- res_bb$caz
noCAZ_bb <- res_bb$nocaz

res_statin <- prepare_med_data(all_statin_data)
CAZ_statin <- res_statin$caz
noCAZ_statin <- res_statin$nocaz
```

### Saving outputs for other analyses

```{r}
#| eval: TRUE
save(
  CAZ_bnf, noCAZ_bnf,
  CAZ_mtlkst, noCAZ_mtlkst,
  CAZ_antd, noCAZ_antd,
  CAZ_anx, noCAZ_anx,
  CAZ_meth, noCAZ_meth,
  CAZ_thiazide, noCAZ_thiazide,
  CAZ_ace, noCAZ_ace,
  CAZ_arb, noCAZ_arb,
  CAZ_bb, noCAZ_bb,
  CAZ_statin, noCAZ_statin,
  noCAZ_SABA, CAZ_SABA,
  CAZ_practices,
  file = "data_raw/trends_data.rda"
)
```

## Extracting trends

### SABA

Aggregating SABA prescription data for control areas (excluding London).

```{r}
noCAZ_SABA_aggre <- noCAZ_SABA |>
  summarise(
    across(c(NUMBER_OF_PATIENTS, numerator, denominator), \(x) {
      sum(x, na.rm = T)
    }),
    .by = c(location_class, month, year, date)
  ) |>
  mutate(calc_value = numerator / denominator)
```

#### SABA ratio

```{r}
plots_SABA_ratio <- CAZ_SABA |>
  left_join(
    CAZ_practices |>
      st_drop_geometry() |>
      select(code, location_class, CAZ_name),
    by = c("org_id" = "code")
  ) |>

  summarise(
    across(c(NUMBER_OF_PATIENTS, numerator, denominator), \(x) {
      sum(x, na.rm = T)
    }),
    .by = c(location_class, CAZ_name, month, year, date)
  ) |>
  mutate(calc_value = numerator / denominator) |>
  nest(.by = c(CAZ_name)) |>
  left_join(CAZ_lst |> select(CAZ, Start), by = c("CAZ_name" = "CAZ")) |>
  mutate(
    trend_data = map2(data, Start, \(.x, .y) {
      .x |>
        bind_rows(noCAZ_SABA_aggre) |>
        filter(year >= .y)
    })
  ) |>
  mutate(
    plots = map2(data, trend_data, \(.x, .y) {
      .x |>
        ggplot(aes(x = date, y = calc_value, colour = location_class)) +
        geom_line(aes(group = location_class), alpha = 0.2, linewidth = 0.2) +
        geom_smooth(
          data = .y,
          method = "lm",
          formula = 'y ~ x',
          se = F,
          aes(linetype = location_class)
        ) +
        theme_minimal() +
        scale_colour_manual(
          values = c(cols4all::c4a(palette = "set1")[1:4], "gray30")
        ) +
        scale_y_continuous(limits = c(0, 0.7)) +
        scale_linetype_manual(values = c(rep("solid", 4), "dashed")) +
        guides(linetype = "none")
    }),
    plots = map2(plots, CAZ_name, \(.x, .y) {
      .x + labs(title = .y, y = "SABA ratio", col = "Location Class", x = "")
    })
  )
```

```{r}
#| message: false
#| error: false
#| warning: false
#| output: false

# Loop to display title and plot
lapply(seq_along(plots_SABA_ratio$plots), \(i) {
  title <- "SABA_ratio"
  caz <- plots_SABA_ratio$CAZ_name[i]
  ggsave(
    plot = plots_SABA_ratio$plots[[i]],
    units = "mm",
    width = 180,
    height = 120,
    dpi = 330,
    filename = file.path(
      here(),
      "qmds_ghpages/figures",
      paste0("trend_", caz, "_", title, ".png")
    )
  )
})
```

#### SABA numerator

```{r}

plots_SABA_numerator <- CAZ_SABA |>
  left_join(
    CAZ_practices |>
      st_drop_geometry() |>
      select(code, location_class, CAZ_name),
    by = c("org_id" = "code")
  ) |>

  summarise(
    across(c(NUMBER_OF_PATIENTS, numerator, denominator), \(x) {
      sum(x, na.rm = T)
    }),
    .by = c(location_class, CAZ_name, month, year, date)
  ) |>
  mutate(calc_value = numerator / denominator) |>
  nest(.by = c(CAZ_name)) |>
  left_join(CAZ_lst |> select(CAZ, Start), by = c("CAZ_name" = "CAZ")) |>
  mutate(
    trend_data = map2(data, Start, \(.x, .y) {
      .x |>
        bind_rows(noCAZ_SABA_aggre) |>
        filter(year >= .y)
    })
  ) |>
  mutate(
    plots = map2(data, trend_data, \(.x, .y) {
      .x |>
        ggplot(aes(
          x = date,
          y = numerator / NUMBER_OF_PATIENTS,
          colour = location_class
        )) +
        geom_line(aes(group = location_class), alpha = 0.2, linewidth = 0.2) +
        geom_smooth(
          data = .y,
          method = "lm",
          formula = 'y ~ x',
          se = F,
          aes(linetype = location_class)
        ) +
        theme_minimal() +
        scale_colour_manual(
          values = c(cols4all::c4a(palette = "set1")[1:4], "gray30")
        ) +
        # scale_y_continuous(limits = c(0,1))+
        scale_linetype_manual(values = c(rep("solid", 4), "dashed")) +
        guides(linetype = "none")
    }),
    plots = map2(plots, CAZ_name, \(.x, .y) {
      .x +
        labs(
          title = .y,
          y = "Prescriptions of SABA inhalers per patient",
          col = "Location Class",
          x = ""
        )
    })
  )
```

```{r}
#| message: false
#| error: false
#| warning: false
#| output: false

# Loop to display title and plot
lapply(seq_along(plots_SABA_numerator$plots), \(i) {
  title <- "SABA_numerator"
  caz <- plots_SABA_numerator$CAZ_name[i]
  ggsave(
    plot = plots_SABA_numerator$plots[[i]],
    units = "mm",
    width = 180,
    height = 120,
    dpi = 330,
    filename = file.path(
      here(),
      "qmds_ghpages/figures",
      paste0("trend_", caz, "_", title, ".png")
    )
  )
})

```

#### SABA denominator (all inhalers)

```{r}
plots_SABA_denominator <- CAZ_SABA |>
  left_join(
    CAZ_practices |>
      st_drop_geometry() |>
      select(code, location_class, CAZ_name),
    by = c("org_id" = "code")
  ) |>

  summarise(
    across(c(NUMBER_OF_PATIENTS, numerator, denominator), \(x) {
      sum(x, na.rm = T)
    }),
    .by = c(location_class, CAZ_name, month, year, date)
  ) |>
  mutate(calc_value = numerator / denominator) |>
  nest(.by = c(CAZ_name)) |>
  left_join(CAZ_lst |> select(CAZ, Start), by = c("CAZ_name" = "CAZ")) |>
  mutate(
    trend_data = map2(data, Start, \(.x, .y) {
      .x |>
        bind_rows(noCAZ_SABA_aggre) |>
        filter(year >= .y)
    })
  ) |>
  mutate(
    plots = map2(data, trend_data, \(.x, .y) {
      .x |>
        ggplot(aes(
          x = date,
          y = denominator / NUMBER_OF_PATIENTS,
          colour = location_class
        )) +
        geom_line(aes(group = location_class), alpha = 0.2, linewidth = 0.2) +
        geom_smooth(
          data = .y,
          method = "lm",
          formula = 'y ~ x',
          se = F,
          aes(linetype = location_class)
        ) +
        theme_minimal() +
        scale_colour_manual(
          values = c(cols4all::c4a(palette = "set1")[1:4], "gray30")
        ) +
        # scale_y_continuous(limits = c(0,1))+
        scale_linetype_manual(values = c(rep("solid", 4), "dashed")) +
        guides(linetype = "none")
    }),
    plots = map2(plots, CAZ_name, \(.x, .y) {
      .x +
        labs(
          title = .y,
          y = "Prescriptions inhalers per patient",
          col = "Location Class",
          x = ""
        )
    })
  )
```

```{r}
#| message: false
#| error: false
#| warning: false
#| output: false

# Loop to display title and plot
lapply(seq_along(plots_SABA_denominator$plots), \(i) {
  title <- "SABA_denominator"
  caz <- plots_SABA_denominator$CAZ_name[i]
  ggsave(
    plot = plots_SABA_denominator$plots[[i]],
    units = "mm",
    width = 180,
    height = 120,
    dpi = 330,
    filename = file.path(
      here(),
      "qmds_ghpages/figures",
      paste0("trend_", caz, "_", title, ".png")
    )
  )
})


```

### Visualisation of other BNF groups and custom indicators

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false

# Define a helper function to aggregate, plot and export trend plots
plot_med_trends <- function(caz_med, nocaz_med) {
  nocaz_aggre <- nocaz_med |>
    summarise(
      across(c(NUMBER_OF_PATIENTS, items, quantity), \(x) sum(x, na.rm = TRUE)),
      .by = c(location_class, month, year, date, bnf)
    )
  nocaz_aggre$date <- as.Date(nocaz_aggre$date)
  caz_med$date <- as.Date(caz_med$date)
  
  nocaz_nest <- nocaz_aggre |>
    nest(.by = bnf, .key = "control_data")
  
  plots_df <- caz_med |>
    left_join(
      CAZ_practices |>
        st_drop_geometry() |>
        select(code, location_class, CAZ_name),
      by = c("row_id" = "code")
    ) |>
    summarise(
      across(c(NUMBER_OF_PATIENTS, items, quantity), \(x) sum(x, na.rm = TRUE)),
      .by = c(location_class, CAZ_name, month, year, date, bnf)
    ) |>
    nest(.by = c(CAZ_name, bnf)) |>
    left_join(
      CAZ_lst |> select(CAZ, Start),
      by = c("CAZ_name" = "CAZ")
    ) |>
    left_join(nocaz_nest, by = "bnf") |>
    mutate(
      trend_data = map2(data, Start, \(.x, .y) {
        .x |> filter(year >= .y)
      }),
      trend_data = map2(trend_data, control_data, \(.x, .y) {
        .x |> bind_rows(.y |> filter(year >= min(.x$year)))
      })
    ) |>
    mutate(
      plots = map2(data, trend_data, \(.x, .y) {
        .x |>
          ggplot(aes(x = date, y = items / NUMBER_OF_PATIENTS, colour = location_class)) +
          geom_line(aes(group = location_class), alpha = 0.2, linewidth = 0.2) +
          geom_smooth(
            data = .y,
            method = "lm",
            formula = 'y ~ x',
            se = FALSE,
            aes(linetype = location_class)
          ) +
          theme_minimal() +
          scale_colour_manual(
            values = c(cols4all::c4a(palette = "set1")[1:4], "gray30")
          ) +
          scale_linetype_manual(values = c(rep("solid", 4), "dashed")) +
          guides(linetype = "none")
      }),
      plots = map2(plots, CAZ_name, \(.x, .y) {
        .x + labs(title = .y, y = "Items per patient", col = "Location Class", x = "")
      }),
      plots = map2(plots, bnf, \(.x, .y) {
        .x + labs(subtitle = paste("BNF code:", .y))
      })
    )
  
  # Save the plots
  lapply(seq_along(plots_df$plots), \(i) {
    title <- plots_df$bnf[i]
    caz <- plots_df$CAZ_name[i]
    ggsave(
      plot = plots_df$plots[[i]],
      units = "mm",
      width = 180,
      height = 120,
      dpi = 330,
      filename = file.path(
        here(),
        "qmds_ghpages/figures",
        paste0("trend_", caz, "_", title, ".png")
      )
    )
  })
  
  plots_df
}

# Run the plotting and export for all indicators
plot_med_trends(CAZ_mtlkst, noCAZ_mtlkst)
plot_med_trends(CAZ_antd, noCAZ_antd)
plot_med_trends(CAZ_anx, noCAZ_anx)
plot_med_trends(CAZ_meth, noCAZ_meth)
plot_med_trends(CAZ_bnf, noCAZ_bnf)

# Run for new cardiovascular indicators
plot_med_trends(CAZ_thiazide, noCAZ_thiazide)
plot_med_trends(CAZ_ace, noCAZ_ace)
plot_med_trends(CAZ_arb, noCAZ_arb)
plot_med_trends(CAZ_bb, noCAZ_bb)
plot_med_trends(CAZ_statin, noCAZ_statin)

```
