---
title: "Age groups and size of practice by boundary types"
---

## Overview

TBC

## Setup

### Package Installation

Loading required packages for data analysis and visualisation.

```{r}
#| label: packages
#| message: false

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("remotes")) {
  install.packages("remotes")
}
pkgs = c(
  "sf",
  "tidyverse",
  "here",
  "tmap",
  "data.table"
)

remotes::install_cran(pkgs)
sapply(pkgs, require, character.only = TRUE)

```

## Loading data

### GP practices

Loading spatial and administrative data for GP practices and CAZ areas.

```{r}
#| label: read-data-pract
subicb_boundaries <- st_read(file.path(
  here(),
  "data_raw",
  "subicb_boundaries.gpkg"
))
pract_classed <- st_read("data_raw/CAZ_practices.gpkg")
all_practices <- st_read(file.path(here(), "data_raw", "gp_locations.gpkg"))
CAZ_lst <- read_csv("data_raw/CAZ_list.csv")
```

### Patient data

Loading patient registration data for calculating prescription rates.

```{r}
#| label: read-data-patients-age

patient_age_df <- read_csv("data_raw/practice_patients_age.csv")

patient_avg_age <- as.data.table(patient_age_df)[,
  `:=`(
    lower = as.numeric(str_extract(AGE_GROUP, "(?<=(\\(|\\[))\\d{1,2}")),
    upper = as.numeric(str_extract(AGE_GROUP, "\\d{2,3}(?=(\\)|\\]))"))
  )
][,
  midpoint := 0.5 * (lower + upper)
][,
  .(mean_age = weighted.mean(midpoint, NUMBER_OF_PATIENTS)),
  by = .(ORG_CODE, month, year)
][ORG_CODE %in% pract_classed$code]

patient_avg_age$month <- match(patient_avg_age$month, tolower(month.name))
```

```{r}
#| label: calculate-practice-size

patient_num_df <- as.data.table(patient_age_df)[,
  .(NUMBER_OF_PATIENTS = sum(NUMBER_OF_PATIENTS)),
  by = .(ORG_CODE, month, year)
][ORG_CODE %in% pract_classed$code]

```

## EDA

### Age distribution 

A quick exploration of the mean age of patients for different CAZ boundaries

```{r}
#| label: map-ages

tmap_mode("view")

pract_classed |>
  inner_join(
    patient_avg_age |>
      filter(year == 2024, month == 12),
    by = c("code" = "ORG_CODE")
  ) |>
  tm_shape() +
  tm_dots("mean_age")

```

```{r}
pract_classed |>
  inner_join(
    patient_avg_age |>
      filter(year == 2024, month == 12),
    by = c("code" = "ORG_CODE")
  ) |>
  ggplot(aes(x = factor(buffer_km), y = mean_age)) +
  geom_jitter(alpha = 0.2, col = "dodgerblue") +
  geom_boxplot(outlier.shape = NA, varwidth = TRUE) +
  facet_wrap(CAZ_name ~ .)
```

#### Variations in mean_age

```{r}
var_mean_age <- patient_avg_age |>
  mutate(
    ym = paste0(year, sprintf("%02d", month)) |> as.numeric(),
    ym_date = ymd(paste0(ym, "01"))
  ) |>
  mutate(baseline = if_else(ym == min(ym), mean_age, NA), .by = ORG_CODE) |>
  mutate(
    baseline = if_else(is.na(baseline), sum(baseline, na.rm = TRUE), baseline),
    .by = ORG_CODE
  ) |>
  mutate(mean_age_std = mean_age - baseline)


ggplot(var_mean_age, aes(mean_age_std)) +
  stat_ecdf() +
  theme_minimal()

```

### Size distribution

```{r}
sf_practices_size <- pract_classed |>
  left_join(
    patient_num_df |>
      summarise(
        across(NUMBER_OF_PATIENTS, \(x) median(x, na.rm = T)),
        .by = ORG_CODE
      ),
    by = c("code" = "ORG_CODE")
  )

```

```{r}
sf_practices_size |>
  ggplot(aes(x = factor(buffer_km), y = NUMBER_OF_PATIENTS)) +
  geom_jitter(alpha = 0.2, col = "dodgerblue") +
  geom_boxplot(outlier.shape = NA, varwidth = TRUE) +
  scale_y_sqrt() +
  facet_wrap(CAZ_name ~ .)
```

### Variation in size

```{r}
patient_num_df$month <- match(patient_num_df$month, tolower(month.name))

var_size_pract <- patient_num_df |>

  mutate(
    ym = paste0(year, sprintf("%02d", month)) |> as.numeric(),
    ym_date = ymd(paste0(ym, "01"))
  ) |>
  mutate(
    baseline = if_else(ym == min(ym), NUMBER_OF_PATIENTS, NA),
    .by = ORG_CODE
  ) |>
  mutate(
    baseline = if_else(is.na(baseline), sum(baseline, na.rm = TRUE), baseline),
    .by = ORG_CODE
  ) |>
  mutate(patients_std = (NUMBER_OF_PATIENTS - baseline) / baseline)


ggplot(var_size_pract, aes(patients_std)) +
  stat_ecdf() +
  theme_minimal()

```