
# Clean Air Zones (CAZ)

## Overview

Clean Air Zones (CAZ) are designated areas within city centers where targeted action is taken to improve air quality by reducing the number of the most polluting vehicles. This section details the geographic boundaries of CAZ areas across the UK and provides access to the spatial data used in this analysis.



## Setup

### Package Installation

Loading required R packages for spatial analysis and data manipulation.

```{r}
#| label: packages
#| message: false

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("remotes")) install.packages("remotes")
pkgs = c(
    "sf",
    "tidyverse",
    "here",
    "tmap"
)

remotes::install_cran(pkgs)
sapply(pkgs, require, character.only = TRUE)
```

## Downloading data

### Boundaries

The boundaries have been obtained manually from the different official repositories and compiled in a
single `gpkg` file. The url to the sources is available in the metadata file.

Downloading CAZ boundary spatial data if not already present locally.

```{r}
#| label: download-boundaries

dir.create("data_raw",showWarnings = F)
if (!file.exists("data_raw/CAZ_boundaries.gpkg")){
  u <- "https://github.com/itsleeds/CAZ-health-data-trends/releases/download/v0/CAZ_boundaries.gpkg"
  f <- basename(u)
  download.file(u,destfile = file.path("data_raw",f),mode = "wb",)
}
```

### Meta-data

Downloading CAZ metadata including implementation dates and zone details.

```{r}
#| label: download-metadata-caz

if (!file.exists("data_raw/CAZ_list.csv")){
  u <- "https://github.com/itsleeds/CAZ-health-data-trends/releases/download/v0/CAZ_list.csv"
  f <- basename(u)
  download.file(u,destfile = file.path("data_raw",f),mode = "wb",)
}
```

## Loading Data

Loading CAZ spatial boundaries and metadata into R for analysis.

```{r}
#| label: load-boundaries-n-metadata

caz_boundaries <- st_read(file.path(here(),"data_raw","CAZ_boundaries.gpkg"), quiet = TRUE)
caz_metadata <- read_csv(file.path(here(),"data_raw","CAZ_list.csv"))
```

## Processing Buffers

Creating buffer zones at different distances around CAZ boundaries for proximity analysis.

```{r}
#| label: make-buffers

buffers <- c(0,0.5,1,5,10)*1e3

caz_buffers <- lapply(buffers, \(buff_dist) {
  st_buffer(caz_boundaries, dist = buff_dist) |>
    mutate(buffer_km = buff_dist / 1e3)
}) |>
  bind_rows()
```

Saving buffer zones to file for use in subsequent analyses.

```{r}
#| label: save-buffers

st_write(caz_buffers,file.path(here(),"data_raw","CAZ_buffers.gpkg"),append = FALSE, quiet = TRUE)
```

A quick visualisation

```{r}
#| label: map-buffers

tmap_mode("view")

tm_shape(caz_buffers |> arrange(-buffer_km))+
  tm_polygons("buffer_km",
              fill_alpha = 0.4,
              fill.scale = tm_scale_discrete(ticks = buffers/1e3,
                                             values = "-brewer.blues")
              )
```
