---
title: "Analysis of Fingertips data"
format: html
---

This file presents an analysis of hospital admissions data for districts with Clean Air Zones (CAZ) in England. The aim is to explore health trends, focusing on hospital admissions of under-19-year-olds, and to visualise geographical and temporal patterns using publicly available datasets.

## Setup

### Package Installation

Loading required packages for data analysis and visualisation.

```{r}
#| label: packages
#| message: false

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("remotes")) {
  install.packages("remotes")
}
pkgs <- c(
  "sf",
  "tidyverse",
  "here",
  "tmap",
  "data.table",
  "plotly",
  "knitr",
  "kableExtra"
)

remotes::install_cran(pkgs)
sapply(pkgs, require, character.only = TRUE)
```

## Loading data

### Hospital Admissions Data

We use hospital admissions data sourced from the Fingertips platform, which provides a wide range of public health indicators. The dataset includes information on admissions by area, age, sex, and time period, allowing for detailed analysis of health outcomes.

```{r}
admissions_data <- read_csv(
  "data_raw/fingertips_data.csv",
  col_types = cols(
    IndicatorID = col_double(),
    IndicatorName = col_character(),
    ParentCode = col_character(),
    ParentName = col_character(),
    AreaCode = col_character(),
    AreaName = col_character(),
    AreaType = col_character(),
    Sex = col_character(),
    Age = col_character(),
    CategoryType = col_logical(),
    Category = col_logical(),
    Timeperiod = col_character(),
    Value = col_double(),
    LowerCI95.0limit = col_double(),
    UpperCI95.0limit = col_double(),
    LowerCI99.8limit = col_double(),
    UpperCI99.8limit = col_double(),
    Count = col_double(),
    Denominator = col_double(),
    Valuenote = col_character(),
    RecentTrend = col_character(),
    ComparedtoEnglandvalueorpercentiles = col_character(),
    ComparedtoParentvalueorpercentiles = col_character(),
    TimeperiodSortable = col_double(),
    Newdata = col_logical(),
    Comparedtogoal = col_logical(),
    Timeperiodrange = col_character()
  )
)
```

### Geographical Data

#### CAZ Boundaries

Spatial boundaries for Clean Air Zones are loaded from a GeoPackage file. These boundaries define the areas where interventions to improve air quality have been implemented.

```{r}
caz_boundaries <- st_read(
  file.path(here(), "data_raw", "CAZ_boundaries.gpkg"),
  quiet = TRUE
)
```

#### Local Authority District Boundaries

Local Authority District (LAD) boundaries are obtained from the ONS GeoPortal and downloaded if not already present. These boundaries are used to identify which districts intersect with CAZ areas.

```{r}
if (!file.exists("data_raw/LAD_MAY_2025_UK_BFC_V2_6634550694215771101.gpkg")) {
  url <- "https://github.com/itsleeds/CAZ-health-data-trends/releases/download/v0/LAD_MAY_2025_UK_BFC_V2_6634550694215771101.gpkg"

  download.file(
    url,
    destfile = "data_raw/LAD_MAY_2025_UK_BFC_V2_6634550694215771101.gpkg",
    mode = "wb"
  )
}

la_boundaries <- st_read(
  "data_raw/LAD_MAY_2025_UK_BFC_V2_6634550694215771101.gpkg",
  quiet = TRUE
)
```

## Pre-processing

### Identifying Districts with CAZ Coverage

To focus the analysis on relevant areas, we identify LADs that overlap with CAZ boundaries. This spatial join ensures that subsequent analyses are restricted to districts affected by CAZ interventions.

```{r}
CAZ_la_boundaries <- la_boundaries[caz_boundaries, ]
```

### Visualising CAZ Locations

We create an interactive map to visualise the location of CAZ districts within England. This helps contextualise the analysis and provides a geographical overview of the study area.

```{r}
tmap_mode("view")

tm_shape(CAZ_la_boundaries) +
  tm_polygons("grey70") +
  tm_shape(caz_boundaries) +
  tm_polygons("blue", fill_alpha = 0.6)
```

### Subsetting Admissions Data

The admissions data is filtered to include only records from districts with CAZ coverage. This step ensures that our health trend analysis is specific to the population impacted by CAZ policies.

```{r}
# Select the indicators we want to plot
sel_indicators <- c(90810, 90986, 212)

# Create a robust lookup mapping for ONS Area Codes to human-readable CAZ names
caz_lookup <- tibble(
  AreaCode = c("E06000022", "E08000025", "E08000032", "E06000023", "E08000021", "E06000044", "E08000019"),
  CAZName = c("BANES", "Birmingham", "Bradford", "Bristol", "Newcastle upon Tyne", "Portsmouth", "Sheffield")
)

# Join the admissions data with our lookup table
CAZ_admissions <- admissions_data |>
  filter(IndicatorID %in% sel_indicators) |>
  inner_join(caz_lookup, by = "AreaCode")
```

## Visualising Trends

### National Control Baseline & Value Normalisation

To compare the CAZ districts against a national trend, we calculate the control group metrics using data from non-CAZ and non-London local authorities. 

Since the selected indicators represent different metrics (emergency hospital admissions per 100,000 for asthma/stroke, vs GP register prevalence percentage for stroke), we dynamically calculate the appropriate multiplier for each indicator from the raw data. This avoids scaling mismatches when drawing control lines.

```{r}
#| label: control-group-prep
#| warning: false
#| message: false

# Define London codes to exclude from the control group
london_codes <- c(
  "E09000006", "E09000023", "E09000022", "E09000008", "E09000011",
  "E09000016", "E09000007", "E09000029", "E09000024", "E09000004",
  "E09000030", "E09000001", "E09000012", "E09000031", "E09000026",
  "E09000025", "E09000010", "E09000019", "E09000033", "E09000003",
  "E09000005", "E09000009", "E09000015", "E09000017", "E09000002",
  "E09000021", "E09000027"
)

# Calculate the appropriate multiplier for each indicator dynamically based on the median ratio
# of Value to (Count / Denominator) in the non-zero raw data
indicator_multipliers <- admissions_data |>
  filter(!is.na(Value), !is.na(Count), !is.na(Denominator), Denominator > 0, Count > 0) |>
  mutate(ratio = Value / (Count / Denominator)) |>
  summarise(
    multiplier = 10^round(log10(median(ratio))),
    .by = IndicatorID
  )

# Calculate control group trends per indicator, time period, and age group using the correct multipliers
admissions_control <- admissions_data |>
  filter(
    IndicatorID %in% sel_indicators,
    !(AreaCode %in% CAZ_la_boundaries$LAD25CD),
    !(AreaCode %in% london_codes),
    Sex == "Persons"
  ) |>
  summarise(
    across(c(Count, Denominator), \(x) sum(x, na.rm = TRUE)),
    .by = c(IndicatorID, TimeperiodSortable, Age)
  ) |>
  left_join(indicator_multipliers, by = "IndicatorID") |>
  mutate(Value = (multiplier * Count) / Denominator, AreaName = "Control") |>
  select(-multiplier)

# Create a lookup vector mapping TimeperiodSortable back to readable financial year strings
timeperiod_labels <- admissions_data |>
  select(TimeperiodSortable, Timeperiod) |>
  distinct() |>
  deframe()
```

### Interactive Trends by Indicator

We plot the trends for each of the selected indicators across all CAZ districts, grouped by age and area. The plots include a dashed black control line representing trends in non-CAZ districts for comparison.

::: {.panel-tabset}

#### Asthma Admissions (Under 19)

**Indicator ID:** 90810 — Hospital admissions for asthma (under 19 years)

This indicator shows the rate of emergency hospital admissions for asthma in children and young people under 19 years of age, expressed per 100,000 population.

```{r}
#| label: plot-asthma-admissions
#| warning: false
#| message: false

df_plot_asthma <- CAZ_admissions |> filter(IndicatorID == 90810, Sex == "Persons")
breaks_asthma <- sort(unique(df_plot_asthma$TimeperiodSortable))

p_asthma <- df_plot_asthma |>
  ggplot(aes(x = TimeperiodSortable, y = Value, col = CAZName)) +
  geom_line(
    data = admissions_control |> filter(IndicatorID == 90810),
    col = "black",
    linetype = "dashed"
  ) +
  geom_line() +
  scale_x_continuous(
    breaks = breaks_asthma,
    labels = \(x) timeperiod_labels[as.character(x)]
  ) +
  facet_grid(Age ~ ., scales = "free_y") +
  theme_light() +
  scale_color_discrete(type = "qual", palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Hospital Admissions for Asthma (under 19 years)",
    y = "per 100,000",
    x = NULL
  )

ggplotly(p_asthma)
```

#### Stroke Admissions

**Indicator ID:** 90986 — Hospital admissions due to stroke

This indicator shows the rate of hospital admissions due to stroke for all ages, expressed per 100,000 population.

```{r}
#| label: plot-stroke-admissions
#| warning: false
#| message: false

df_plot_stroke <- CAZ_admissions |> filter(IndicatorID == 90986, Sex == "Persons")
breaks_stroke <- sort(unique(df_plot_stroke$TimeperiodSortable))

p_stroke <- df_plot_stroke |>
  ggplot(aes(x = TimeperiodSortable, y = Value, col = CAZName)) +
  geom_line(
    data = admissions_control |> filter(IndicatorID == 90986),
    col = "black",
    linetype = "dashed"
  ) +
  geom_line() +
  scale_x_continuous(
    breaks = breaks_stroke,
    labels = \(x) timeperiod_labels[as.character(x)]
  ) +
  facet_grid(Age ~ ., scales = "free_y") +
  theme_light() +
  scale_color_discrete(type = "qual", palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Hospital Admissions due to Stroke",
    y = "per 100,000",
    x = NULL
  )

ggplotly(p_stroke)
```

#### Stroke Prevalence (QOF)

**Indicator ID:** 212 — Stroke: QOF prevalence

This indicator shows the percentage of patients on the GP register with a diagnosis of stroke or TIA (Quality and Outcomes Framework), representing overall prevalence in the population.

```{r}
#| label: plot-stroke-prevalence
#| warning: false
#| message: false

df_plot_prev <- CAZ_admissions |> filter(IndicatorID == 212, Sex == "Persons")
breaks_prev <- sort(unique(df_plot_prev$TimeperiodSortable))

p_prev <- df_plot_prev |>
  ggplot(aes(x = TimeperiodSortable, y = Value, col = CAZName)) +
  geom_line(
    data = admissions_control |> filter(IndicatorID == 212),
    col = "black",
    linetype = "dashed"
  ) +
  geom_line() +
  scale_x_continuous(
    breaks = breaks_prev,
    labels = \(x) timeperiod_labels[as.character(x)]
  ) +
  facet_grid(Age ~ ., scales = "free_y") +
  theme_light() +
  scale_color_discrete(type = "qual", palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Stroke: QOF Prevalence",
    y = "Prevalence (%)",
    x = NULL
  )

ggplotly(p_prev)
```

:::

### Data Table

```{r}
#| label: data-table
#| warning: false

CAZ_admissions |>
  filter(Sex == "Persons") |>
  knitr::kable(format = "html") |>
  kable_paper() |>
  scroll_box(width = "100%", height = "500px")
```

### Exporting Plots for Individual CAZ Report Pages

To support the individual Clean Air Zone report pages (which render static PNG plots for comparison), we save plots for each CAZ. 

1. `admissions_<caz_name>.png` - Static plot for the primary indicator (90810: Asthma admissions under 19 years), as expected by the results templates.
2. `admissions_<ind>_<caz_name>.png` - Static plots for each individual indicator and CAZ.

```{r}
#| label: export-plots
#| warning: false
#| message: false

# Ensure the figures directory exists
dir.create("qmds_ghpages/figures", showWarnings = FALSE, recursive = TRUE)

# Loop over each CAZ to save individual plots
for (caz_row in seq_len(nrow(caz_lookup))) {
  caz_code <- caz_lookup$AreaCode[caz_row]
  caz_name <- caz_lookup$CAZName[caz_row]
  
  # 1. Save the primary indicator plot (90810) as admissions_<caz_name>.png for the results pages
  ind_primary <- 90810
  ind_name_primary <- admissions_data |>
    filter(IndicatorID == ind_primary) |>
    pull(IndicatorName) |>
    unique()
  
  df_plot_prim_caz <- CAZ_admissions |> 
    filter(IndicatorID == ind_primary, AreaCode == caz_code, Sex == "Persons")
  breaks_prim_caz <- sort(unique(df_plot_prim_caz$TimeperiodSortable))
  
  p_primary <- df_plot_prim_caz |>
    ggplot(aes(x = TimeperiodSortable, y = Value)) +
    geom_line(
      data = admissions_control |> filter(IndicatorID == ind_primary),
      col = "black",
      linetype = "dashed"
    ) +
    geom_line(aes(col = CAZName)) +
    scale_x_continuous(
      breaks = breaks_prim_caz,
      labels = \(x) timeperiod_labels[as.character(x)]
    ) +
    facet_grid(Age ~ ., scales = "free_y") +
    theme_light() +
    scale_color_discrete(type = "qual", palette = "Dark2") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    guides(col = "none") +
    labs(
      title = paste0(ind_name_primary, " — ", caz_name),
      y = "per 100,000",
      x = NULL
    )
  
  ggsave(
    filename = paste0("qmds_ghpages/figures/admissions_", caz_name, ".png"),
    plot = p_primary,
    width = 15,
    height = 10,
    dpi = 300,
    units = "cm"
  )
  
  # 2. Save individual plots for all indicators (including 90810, 90986, 212)
  for (ind in sel_indicators) {
    ind_name <- admissions_data |>
      filter(IndicatorID == ind) |>
      pull(IndicatorName) |>
      unique()
    
    y_label <- ifelse(ind == 212, "Prevalence (%)", "per 100,000")
    
    df_plot_ind_caz <- CAZ_admissions |> 
      filter(IndicatorID == ind, AreaCode == caz_code, Sex == "Persons")
    breaks_ind_caz <- sort(unique(df_plot_ind_caz$TimeperiodSortable))
    
    p_ind <- df_plot_ind_caz |>
      ggplot(aes(x = TimeperiodSortable, y = Value)) +
      geom_line(
        data = admissions_control |> filter(IndicatorID == ind),
        col = "black",
        linetype = "dashed"
      ) +
      geom_line(aes(col = CAZName)) +
      scale_x_continuous(
        breaks = breaks_ind_caz,
        labels = \(x) timeperiod_labels[as.character(x)]
      ) +
      facet_grid(Age ~ ., scales = "free_y") +
      theme_light() +
      scale_color_discrete(type = "qual", palette = "Dark2") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      guides(col = "none") +
      labs(
        title = paste0(ind_name, " — ", caz_name),
        y = y_label,
        x = NULL
      )
    
    caz_filename_clean <- gsub(" ", "_", caz_name)
    ggsave(
      filename = paste0("qmds_ghpages/figures/admissions_", ind, "_", caz_filename_clean, ".png"),
      plot = p_ind,
      width = 15,
      height = 10,
      dpi = 300,
      units = "cm"
    )
  }
}
```
