---
title: "prescription_data"
format: html
---

## Overview

This section handles the acquisition and processing of prescription data from NHS England, focusing particularly on medications relevant to respiratory health. The primary focus is on Short-Acting Beta Agonists (SABA), which are commonly prescribed rescue inhalers for asthma and other respiratory conditions.

The prescription data is sourced from OpenPrescribing.net, which provides monthly prescribing data for all NHS practices in England. This allows us to:

- Track prescription trends over time across different geographic areas
- Compare prescribing patterns between CAZ and non-CAZ areas
- Adjust for population size using patient registration data
- Identify potential changes in respiratory health following CAZ implementation

This analysis forms the core of our investigation into the potential health impacts of Clean Air Zones.

## Setup

### Package Installation

Loading required packages for spatial analysis and web scraping.

```{r}
#| label: packages
#| message: false

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("remotes")) {
  install.packages("remotes")
}
pkgs <- c(
  "sf",
  "tidyverse",
  "here",
  "tmap",
  "geojsonsf",
  "rvest",
  "httr"
)

remotes::install_cran(pkgs)
sapply(pkgs, require, character.only = TRUE)
```

## Loading data

Loading Sub-ICB boundary data for geographic reference.

```{r}
#| label: load-boundaries

subicb_boundaries <- st_read(file.path(
  here(),
  "data_raw",
  "subicb_boundaries.gpkg"
), quiet = TRUE)
```

### SABA

Downloading SABA prescription data from OpenPrescribing for all Sub-ICB areas.

```{r}
#| label: download-saba
#| message: false

if (!file.exists(file.path(here(), "data_raw", "all_saba.csv"))) {
  all_saba_raw <- do.call(
    bind_rows,
    lapply(subicb_boundaries$code, \(t_code) {
      read_csv(
        paste0(
          "https://openprescribing.net/api/1.0/measure_by_practice/?format=csv&org=",
          t_code,
          "&parent_org_type=ccg&measure=saba"
        ),
        col_types = cols(
          measure = col_character(),
          org_type = col_character(),
          org_id = col_character(),
          org_name = col_character(),
          date = col_date(format = ""),
          numerator = col_double(),
          denominator = col_double(),
          calc_value = col_double(),
          percentile = col_double()
        )
      )
    })
  )
  write_csv(all_saba_raw, file = file.path(here(), "data_raw", "all_saba.csv"))
} else {
  all_saba_raw <- read_csv(file = file.path(here(), "data_raw", "all_saba.csv"))
}
```


## All BNF groups

### BNF Code Discovery

Extracting all available BNF codes from the OpenPrescribing website for comprehensive data download.

```{r}
#| label: download-all

base_u <- "https://openprescribing.net/bnf/"
bnf_site <- read_html(base_u)

links <- bnf_site |> html_nodes("a") |> html_attr("href")
links <- links[grep("/bnf/\\d{2,6}/$", links)] |>
  str_remove("/bnf/") |>
  str_remove("/$")
links <- links[order(nchar(links))]

dates <- seq(as.Date("2021-05-01"), as.Date("2026-04-01"), "months")

full_data_grid <- expand_grid(code = links, date = dates) |>
  mutate(group = nchar(code), file.name = paste0(code, "_", date, ".csv"))

write_rds(full_data_grid, file = "data_raw/full_bnf_grid.rds")

# Helper function to read from the OpenPrescribing API with browser headers to avoid Cloudflare blocks
read_openprescribing <- function(url) {
  resp <- GET(url,
              add_headers(Accept = "application/json"),
              user_agent("R httr"))
  stop_for_status(resp)
  content(resp, as = "text", encoding = "UTF-8")
}

# Define a robust download helper function with rate-limiting and error handling
download_bnf_grid <- function(grid_df) {
  dir.create("data_raw/bnf_all", showWarnings = FALSE)
  for (i in seq_len(nrow(grid_df))) {
    dest <- file.path("data_raw", "bnf_all", grid_df$file.name[i])
    if (!file.exists(dest)) {
      url <- paste0(
        "https://openprescribing.net/api/1.0/spending_by_org/?org_type=practice&code=",
        grid_df$code[i],
        "&date=",
        grid_df$date[i],
        "&format=csv"
      )
      tryCatch({
        csv_text <- read_openprescribing(url)
        
        # Verify response looks like a CSV (not empty and not HTML)
        if (nchar(csv_text) > 50 && !grepl("<!DOCTYPE html>", csv_text, fixed = TRUE)) {
          writeLines(csv_text, dest, useBytes = TRUE)
          Sys.sleep(0.1) # be considerate to the API rate limit
        } else {
          warning(paste("Invalid file content returned for:", url))
        }
      }, error = function(e) {
        warning(paste("Failed to download:", url, "-", e$message))
        if (file.exists(dest)) file.remove(dest)
      })
    }
  }
}

all_data_group2 <- full_data_grid |> filter(group == 2)
download_bnf_grid(all_data_group2)
```

### Montelukast

```{r}
#| label: download-montelukast
#| warning: false
#| message: false

mont_data_grid <- expand_grid(code = "0303020G0", date = dates) |>
  mutate(group = nchar(code), file.name = paste0(code, "_", date, ".csv"))

download_bnf_grid(mont_data_grid)
```

### Anxiolytics and Antidepressants

```{r}
#| label: download-anxiolytics-antidepressants
#| warning: false
#| message: false

mont_data_grid_anx_antd <- expand_grid(code = c("0403", "040102"), date = dates) |>
  mutate(group = nchar(code), file.name = paste0(code, "_", date, ".csv"))

download_bnf_grid(mont_data_grid_anx_antd)
```

### Oral Steroid Methylprednisolone

```{r}
#| label: download-methylprednisolone
#| warning: false
#| message: false

meth_data_grid <- expand_grid(code = "0603020S0", date = dates) |>
  mutate(group = nchar(code), file.name = paste0(code, "_", date, ".csv"))

download_bnf_grid(meth_data_grid)
```

### Cardiovascular Indicators

```{r}
#| label: download-cardio-indicators
#| warning: false
#| message: false

cardio_codes <- c("020201", "0205051", "020505", "0204", "0212")
cardio_grid <- expand_grid(code = cardio_codes, date = dates) |>
  mutate(group = nchar(code), file.name = paste0(code, "_", date, ".csv"))

download_bnf_grid(cardio_grid)
```
