---
title: "Difference-in-Differences Analysis"
params:
  current_caz: "Bradford"
format: html
execute:
  warning: false
  message: false
---

:::{.callout-warning}
These are preliminary results and the analysis is still in progress.
:::

## Overview

This document implements a Difference-in-Differences (DiD) analysis to estimate the causal effect of **`{r} params$current_caz`**'s Clean Air Zone (CAZ) implementation on total inhaler prescriptions (SABA denominator). The DiD approach compares changes in prescription rates before and after CAZ implementation in areas within 10 km of the CAZ (treatment group) relative to changes in control areas (practices in Major Towns and cities outside any CAZ excluding London).

## Setup

```{r}
#| label: packages

library(tidyverse)
library(here)
library(broom)
library(sf)
library(knitr)
library(kableExtra)


```

```{r}
#| label: load-data

load(file.path(here(), "data_raw/trends_data.rda"))

CAZ_lst <- read_csv(
  file.path(here(), "data_raw/CAZ_list.csv"),
  show_col_types = FALSE
)
```

## Data Preparation for `{r} params$current_caz`

```{r}
#| label: prep-data

# Get CAZ start date
caz_start_year <- CAZ_lst |>
  filter(CAZ == params$current_caz) |>
  pull(Start)

caz_start <- as.Date(paste0(caz_start_year, "-01-01"))

# Filter CAZ practices
caz_saba <- CAZ_SABA |>
  left_join(
    CAZ_practices |>
      st_drop_geometry() |>
      select(code, location_class, CAZ_name),
    by = c("org_id" = "code")
  ) |>
  filter(CAZ_name == params$current_caz)

# Aggregate CAZ data
caz_agg <- caz_saba |>
  filter(str_detect(location_class, "within")) |>
  summarise(
    total_patients = sum(NUMBER_OF_PATIENTS, na.rm = TRUE),
    total_numerator = sum(numerator, na.rm = TRUE),
    total_denominator = sum(denominator, na.rm = TRUE),
    .by = c(location_class, date, month, year)
  ) |>
  filter(total_patients > 0) |>
  mutate(
    inhalers_per_1000 = (total_denominator / total_patients) * 1000,
    treatment = 1
  )

# Aggregate control data
control_agg <- noCAZ_SABA |>
  summarise(
    total_patients = sum(NUMBER_OF_PATIENTS, na.rm = TRUE),
    total_numerator = sum(numerator, na.rm = TRUE),
    total_denominator = sum(denominator, na.rm = TRUE),
    .by = c(date, month, year)
  ) |>
  filter(total_patients > 0) |>
  mutate(
    inhalers_per_1000 = (total_denominator / total_patients) * 1000,
    treatment = 0,
    location_class = "Control"
  )

# Combine treatment and control
did_data <- bind_rows(
  caz_agg,
  control_agg
) |>
  mutate(
    post = as.integer(date >= caz_start),
    time_to_treatment = as.numeric(difftime(date, caz_start, units = "days")) /
      30.44 # Convert to months
  ) |>
  filter(
    !is.na(inhalers_per_1000),
    is.finite(inhalers_per_1000)
  ) |>
  arrange(date, treatment)
```

## Summary Statistics

```{r}
#| label: summary-stats

did_summary <- did_data |>
  group_by(treatment, post) |>
  summarise(
    n_obs = n(),
    mean_inhalers = mean(inhalers_per_1000, na.rm = TRUE),
    sd_inhalers = sd(inhalers_per_1000, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    period = ifelse(post == 0, "Pre-CAZ", "Post-CAZ"),
    group = ifelse(treatment == 1, params$current_caz, "Control")
  )

kable(
  did_summary |>
    select(group, period, n_obs, mean_inhalers, sd_inhalers),
  digits = 2,
  col.names = c("Group", "Period", "N Obs", "Mean Inhalers/1000", "SD"),
  caption = "Summary Statistics by Treatment and Period"
) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Parallel Trends Visualization

```{r}
#| label: fig-parallel-trends
#| fig-width: 10
#| fig-height: 6

plot_data <- did_data |>
  group_by(date, treatment) |>
  summarise(
    mean_inhalers = mean(inhalers_per_1000, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    group = ifelse(treatment == 1, "CAZ", "Control")
  )

ggplot(
  plot_data,
  aes(x = date, y = mean_inhalers, color = group, linetype = group)
) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = caz_start,
    linetype = "dashed",
    color = "red",
    linewidth = 0.8
  ) +
  annotate(
    "text",
    x = caz_start,
    y = max(plot_data$mean_inhalers, na.rm = TRUE),
    label = "CAZ Implementation",
    vjust = -0.5,
    color = "red"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  ) +
  labs(
    title = paste(
      "Total Inhaler Prescriptions:",
      "CAZ",
      "vs Control"
    ),
    subtitle = "Per 1,000 registered patients",
    x = "Date",
    y = "Inhalers per 1,000 patients",
    color = "Group",
    linetype = "Group"
  ) +
  scale_color_manual(
    values = c("CAZ" = "#E69F00", "Control" = "#0072B2")
  )
```

## Difference Plot

Difference plot with 12-month rolling average to smooth out short-term fluctuations and highlight overall trends.

```{r}
#| label: fig-did-raw-difference
#| fig-width: 10
#| fig-height: 6

diff_plot_data <- did_data |>
  group_by(date, treatment) |>
  summarise(
    mean_inhalers = mean(inhalers_per_1000, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    group = ifelse(treatment == 1, "CAZ", "Control")
  ) |>
  select(-treatment) |>
  pivot_wider(names_from = group, values_from = mean_inhalers) |>
  mutate(
    caz_rm = zoo::rollmean(
      get("CAZ"),
      k = 12,
      fill = NA,
      align = "right"
    ),
    control_rm = zoo::rollmean(Control, k = 12, fill = NA, align = "right")
  ) |>
  mutate(
    difference_rm = caz_rm - control_rm,
    difference = get("CAZ") - Control
  )

ggplot(diff_plot_data, aes(x = date)) +
  geom_line(
    aes(y = difference, color = "Monthly Difference"),
    linewidth = 0.5
  ) +
  geom_line(
    aes(y = difference_rm, color = "12-month Rolling Average"),
    linewidth = 1
  ) +
  geom_vline(
    xintercept = caz_start,
    linetype = "dashed",
    color = "red",
    linewidth = 0.8
  ) +
  annotate(
    "text",
    x = caz_start,
    y = max(diff_plot_data$difference, na.rm = TRUE),
    label = "CAZ Implementation",
    vjust = -0.5,
    hjust = -0.1,
    color = "red"
  ) +
  labs(
    title = paste(
      "Difference in Inhaler Prescriptions:",
      "CAZ",
      "vs Control"
    ),
    subtitle = "Per 1,000 registered patients",
    x = "Date",
    y = "Inhalers per 1,000 patients"
  ) +
  theme_minimal(base_size = 12) +
  scale_y_continuous(expand = 0.2) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold")
  ) +
  scale_color_manual(
    values = c(
      "Monthly Difference" = "#888888",
      "12-month Rolling Average" = "black"
    ),
    name = ""
  )
```

## Difference-in-Differences Models

### Basic DiD Model

```{r}
#| label: did-basic

did_model_basic <- lm(
  inhalers_per_1000 ~ treatment * post,
  data = did_data
)

summary(did_model_basic)

did_coef_basic <- tidy(did_model_basic, conf.int = TRUE) |>
  filter(term == "treatment:post")

kable(
  did_coef_basic,
  digits = 3,
  caption = "DiD Estimate (Treatment × Post interaction)"
) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### DiD with Time Trends

```{r}
#| label: did-trends

did_model_trends <- lm(
  inhalers_per_1000 ~ treatment * post + time_to_treatment,
  data = did_data
)

summary(did_model_trends)

did_coef_trends <- tidy(did_model_trends, conf.int = TRUE) |>
  filter(term == "treatment:post")

kable(
  did_coef_trends,
  digits = 3,
  caption = "DiD Estimate with Time Trends (Treatment × Post interaction)"
) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### DiD with Month Fixed Effects

```{r}
#| label: did-month-fe

did_model_month <- lm(
  inhalers_per_1000 ~ treatment * post + factor(month),
  data = did_data
)

summary(did_model_month)

did_coef_month <- tidy(did_model_month, conf.int = TRUE) |>
  filter(term == "treatment:post")

kable(
  did_coef_month,
  digits = 3,
  caption = "DiD Estimate with Month Fixed Effects (Treatment × Post interaction)"
) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Comparison Across Specifications

```{r}
#| label: comparison-table

comparison <- bind_rows(
  did_coef_basic |> mutate(model = "Basic DiD"),
  did_coef_trends |> mutate(model = "With Time Trend"),
  did_coef_month |> mutate(model = "With Month FE")
) |>
  select(model, estimate, std.error, statistic, p.value, conf.low, conf.high)

kable(
  comparison,
  digits = 3,
  caption = "Comparison of DiD Estimates Across Specifications"
) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```


## Interpretation

**DiD Estimate:** `{r} sprintf("%.3f", did_coef_basic$estimate)` inhalers per 1,000 patients (95% CI: [`{r} sprintf("%.3f", did_coef_basic$conf.low)`, `{r} sprintf("%.3f", did_coef_basic$conf.high)`])

**P-value:** `{r} sprintf("%.3f", did_coef_basic$p.value)`

**Statistical Significance:** `{r} ifelse(did_coef_basic$p.value < 0.05, "Statistically significant at 5% level", "Not statistically significant at 5% level")`

### Key Findings

1. **Direction**: The estimate indicates a `{r} ifelse(did_coef_basic$estimate > 0, "positive", "negative")` effect on inhaler prescriptions

2. **Magnitude**: The CAZ is associated with a change of approximately `{r} abs(round(did_coef_basic$estimate, 2))` inhalers per 1,000 patients

3. **Robustness**: The estimate is `{r} ifelse(abs(max(comparison$estimate) - min(comparison$estimate)) < 0.5, "stable", "somewhat variable")` across different model specifications

---

*Results produced on: `{r} Sys.Date()`*
